Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds

### 2020-08-29 Manuscript_Publication.Rmd
updated supp info

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds
author:
bibliography: References/SS_Extinction.bib
output:
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{draft, colorlinks=false, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. **Theoretical work also suggests that the net effect of sexual selection on diversification should depend strongly on ecological factors, though this prediction has seldom been tested. Here, we test whether variation in sexual selection can predict speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera) and whether this relationship is mediated by environmental factors. Male-biased sexual selection, and specifically sexual size dimorphism, predicted two of the three measures of speciation rates that we examined**. The link we observed between sexual selection and speciation was independent of environmental variability, though species with smaller ranges had higher speciation rates. There was no association between any proxies of sexual selection and extinction rate. **Our findings support the view that male-biased sexual selection, as measured by predictors of male-male competition, has shaped diversification in the largest radiation of birds.**
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

@@

We examined the effect of sexual selection on speciation and extinction rate in 97% of passerines (*n* = 5,812 species; 58% of all birds). Specifically, we (*i*) compiled datasets for sexual dichromatism/selection strength and environmental variability, (*ii*) obtained estimates of speciation and extinction rates across passerines, and (*iii*) conducted phylogenetic generalized least-squares (PGLS) regressions. Analyses are documented with reproducible code in the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}.  \phantomsection

@@

Additionally, we used another measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from museum specimens as opposed to illustrations. The spectrophotometry data covers only 581 passerine species (10-fold fewer than the RGB data), although there was a substantial correlation between the two dichromatism measures ($r = 0.79$; \href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

@@

Because sexual dichromatism is presumably imperfectly correlated with variation in the strength of sexual selection across taxa, we sourced an additional measure of sexual selection [@Dale_2015], referred to here as the 'index of male-biased sexual selection'. This index is the first principal component from a phylogenetic principal component analysis (PPCA) of three characteristics possitively associated with sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care). **The variables included in this index have all been positively linked to the intensity of sexual selection, and are usually correlated [@bjorklund_1990; @owens_1998; @dunn_2001], which is why they were combined into a single metric in previous studies [@Dale_2015]**. This measure of male-biased sexual selection is available for only 2,465 species, and shows a moderate correlation with the RGB measure of sexual dichromatism ($r = 0.34$; \href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Figure S12}).  \phantomsection

@@

We obtained estimates of species range size using expert range maps [@birdlife_2017]. Because of taxonomic changes to 1,230 species in the Birdlife database [@delhoyo_2016] we manually matched these taxa with the names used in the sexual dichromatism dataset [@Dale_2015]. From these distributions, we obtained estimates of climatic conditions that each species is exposed to. That is, we took 1,000 random point samples in each range and extracted 19 present-day bioclimatic variables (representing a variety of biologically relevant annual trends in temperature and precipitation) with 30-second (~1 km^2^) spatial resolution [@Fick_2017]. From these values, we obtained means and standard deviations for each species. Using the same spatial sampling, we extracted means and standard deviations of bioclimatic variables from the paleoclimate during the last interglacial (LIG; 120,000 - 140,000 years ago) [@Otto-Bliesner_2006]. To estimate variability in the energy available to species, we obtained the mean and standard deviation of net primary productivity (NPP) values between 2000 - 2015 across each species distribution. Estimates of NPP had 30-second resolution and were obtained through MODIS (Moderate Resolution Imaging Spectroradiometer) primary production products stage 3 (MOD17A3) [@Zhao_2005]. Using these data, which we provide as a potentially useful data resource (see \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}), we generated five predictors of speciation associated with different patterns in environmental variability (see below).  \phantomsection

@@

Given that stressful environments are expected to interact with sexual selection and have a positive effect on adaptation [@Cally_2019], we used the extracted environmental variables from each species range size to create predictors of environmental variation/stress. Firstly we used (*i*) the average NPP in each species' range and (*ii*) the log-transformed range size as potentially informative predictors of speciation rates. We also used three environmental predictors derived from bioclimatic data. These predictors relate to seasonal climate variation, spatial climate variation and long-term climate variation. To obtain seasonal climate variation we used (*iii*) mean values of temperature seasonality (BIO4) for each range. (*iv*) To estimate levels of spatial environmental variation a species may endure, we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature and precipitation seasonality (BIO4 and BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 **largely** reflects the variation in temperature across a species' range (\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). **A taxon’s range size often correlates with speciation and extinction rates [@Rosenzweig_1995; @Castiglione_2017]**, so we controlled for the correlation between environmental spatial variation and species' range sizes --- where larger ranges have larger variation in PC1 --- by using the residuals of a fitted general additive model (GAM; \href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}) as a predictor. To obtain long-term variation in climates for each species range we take (*v*) the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}). The five predictors of environmental variability are not strongly correlated (\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and ``R`` code to generate these predictors can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}.  \phantomsection

@@

We used BAMM to model the dynamics of speciation and extinction across the **101 phylogenetic trees (one MCC tree and 100 random draws of the posterior)**. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates [@Rabosky_2014]. **These models provide tip-rate estimates of speciation and extinction rate that can be easily used in comparative analyses**. The parameters of the 100 BAMM runs are detailed in full in the \href{https://justincally.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{Supplementary Information}; **briefly, we used a time-variable model with the prior expected number of evolutionary rate shifts set at 100 and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. BAMM models were run independently for the 101 phylogenetic trees for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Important BAMM parameters (log-likelihood and number of rate shifts) reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200; an arbitrary value, above which posterior distributions can often be accurately inferred** (\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.  \phantomsection

@@

Finally, using the subset of species with an index of male-biased sexual selection, we conducted a phylogenetic path analysis using the ``phylopath R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths between variables unable to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection, and range size to have effects on each other and not just on speciation rate. To minimise path complexity we used temperature seasonality (BIO4) as the single measure for environmental variability, $\lambda_{DR}$ as the single measure of speciation and the tip-rates from the MCC tree. Further details of the path analysis, including our rationale for each path's directions, can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information} along with all other analyses and the relevant ``R`` code to reproduce results.  \phantomsection

@@

We found a significant positive association between the index of male-biased sexual selection (*n* = 2,465) and $\lambda_{DR}$ maximum credibility (MCC) tree ($\beta$ = `r sanitize.numbers(formatC(Subset.MCC.DR.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.DR.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; \autoref{model_results}b). However, this association was not significant for the other two measures of speciation rate ($\lambda_{ND}$: $\beta$ = `r sanitize.numbers(formatC(Subset.MCC.ND.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.ND.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; $\lambda_{BAMM}$: $\beta$ = `r sanitize.numbers(formatC(Subset.MCC.Lambda.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.Lambda.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; \autoref{model_results}b). When we took into account phylogenetic uncertainty by running the models using 1,000 trees, the distribution of estimates from PGLS models was similar to the estimate from the MCC tree: among the 1,000 trees there was a positive association between sexual selection and $\lambda_{DR}$ (highest posterior density (HPD) Interval = `r sanitize.numbers(formatC(Subset.hpd.DR.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`), and the distribution skewed towards a positive association between sexual selection and $\lambda_{ND}$ (HPD Interval = `r sanitize.numbers(formatC(Subset.hpd.ND.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`; ) as well as the 100 models using $\lambda_{BAMM}$ (HPD Interval = `r sanitize.numbers(formatC(Subset.hpd.Lambda.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`; \href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Table S15}).

@@

In contrast to male-biased sexual selection, we found no evidence that species with increased sexual dichromatism have higher or lower rates of speciation. Sexual dichromatism showed no association with $\lambda_{DR}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.DR.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.DR.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a, \autoref{Speciation_Figure}), $\lambda_{ND}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.ND.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.ND.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a) or $\lambda_{BAMM}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.Lambda.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.Lambda.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a). PGLS analyses using sexual dichromatism (*n* = 581) measured by spectrophotometry [@Armenta_2008] yielded results concordant with the full dataset; i.e. no association between sexual dichromatism and speciation  (\href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S11}). Our results from models based on the MCC tree are largely corroborated by model estimates from PGLS analyses of the rates and correlation structures from 1,000 trees (for $\lambda_{DR}$, $\lambda_{ND}$) and 100 trees for $\lambda_{BAMM}$. The HPD intervals show model estimates are distributed around zero when using complete taxon sampling models and RGB measures of sexual dichromatism ($\lambda_{DR}$: HPD Interval = `r sanitize.numbers(formatC(hpd.DR.top$SDi, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, $\lambda_{ND}$: HPD Interval = `r sanitize.numbers(formatC(hpd.ND.top$SDi, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, \autoref{model_results}a, \href{https://justincally.github.io/SexualSelection_Speciation/#pgls_model_summary}{Table S8}). For PGLS models using spectrophotometry-based measures of sexual dichromatism, the estimates from the 100 trees in the $\lambda_{DR}$ models are positively skewed (HPD Interval = `r sanitize.numbers(formatC(Armenta.hpd.DR.top$Colour.discriminability, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`) but normally distributed around zero for $\lambda_{ND}$ and $\lambda_{BAMM}$ (\href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Table S12}).

>>>>

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds
author:
bibliography: References/SS_Extinction.bib
output:
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. **Theoretical work also suggests that the net effect of sexual selection on diversification should depend strongly on ecological factors, though this prediction has seldom been tested. Here, we test whether variation in sexual selection can predict speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera) and whether this relationship is mediated by environmental factors. Male-biased sexual selection, and specifically sexual size dimorphism, predicted two of the three measures of speciation rates that we examined**. The link we observed between sexual selection and speciation was independent of environmental variability, though species with smaller ranges had higher speciation rates. There was no association between any proxies of sexual selection and extinction rate. **Our findings support the view that male-biased sexual selection, as measured by predictors of male-male competition, has shaped diversification in the largest radiation of birds.**
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

@@

We examined the effect of sexual selection on speciation and extinction rate in 97% of passerines (*n* = 5,812 species; 58% of all birds). Specifically, we (*i*) compiled datasets for sexual dichromatism/selection strength and environmental variability, (*ii*) obtained estimates of speciation and extinction rates across passerines, and (*iii*) conducted phylogenetic generalized least-squares (PGLS) regressions. Analyses are documented with reproducible code in the \href{https://ssspeciation.github.io/SexualSelection_Speciation/}{Supplementary Information}.  \phantomsection

@@

Additionally, we used another measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from museum specimens as opposed to illustrations. The spectrophotometry data covers only 581 passerine species (10-fold fewer than the RGB data), although there was a substantial correlation between the two dichromatism measures ($r = 0.79$; \href{https://ssspeciation.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

@@

Because sexual dichromatism is presumably imperfectly correlated with variation in the strength of sexual selection across taxa, we sourced an additional measure of sexual selection [@Dale_2015], referred to here as the 'index of male-biased sexual selection'. This index is the first principal component from a phylogenetic principal component analysis (PPCA) of three characteristics possitively associated with sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care). **The variables included in this index have all been positively linked to the intensity of sexual selection, and are usually correlated [@bjorklund_1990; @owens_1998; @dunn_2001], which is why they were combined into a single metric in previous studies [@Dale_2015]**. This measure of male-biased sexual selection is available for only 2,465 species, and shows a moderate correlation with the RGB measure of sexual dichromatism ($r = 0.34$; \href{https://ssspeciation.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Figure S12}).  \phantomsection

@@

We obtained estimates of species range size using expert range maps [@birdlife_2017]. Because of taxonomic changes to 1,230 species in the Birdlife database [@delhoyo_2016] we manually matched these taxa with the names used in the sexual dichromatism dataset [@Dale_2015]. From these distributions, we obtained estimates of climatic conditions that each species is exposed to. That is, we took 1,000 random point samples in each range and extracted 19 present-day bioclimatic variables (representing a variety of biologically relevant annual trends in temperature and precipitation) with 30-second (~1 km^2^) spatial resolution [@Fick_2017]. From these values, we obtained means and standard deviations for each species. Using the same spatial sampling, we extracted means and standard deviations of bioclimatic variables from the paleoclimate during the last interglacial (LIG; 120,000 - 140,000 years ago) [@Otto-Bliesner_2006]. To estimate variability in the energy available to species, we obtained the mean and standard deviation of net primary productivity (NPP) values between 2000 - 2015 across each species distribution. Estimates of NPP had 30-second resolution and were obtained through MODIS (Moderate Resolution Imaging Spectroradiometer) primary production products stage 3 (MOD17A3) [@Zhao_2005]. Using these data, which we provide as a potentially useful data resource (see \href{https://ssspeciation.github.io/SexualSelection_Speciation/}{Supplementary Information}), we generated five predictors of speciation associated with different patterns in environmental variability (see below).  \phantomsection

@@

Given that stressful environments are expected to interact with sexual selection and have a positive effect on adaptation [@Cally_2019], we used the extracted environmental variables from each species range size to create predictors of environmental variation/stress. Firstly we used (*i*) the average NPP in each species' range and (*ii*) the log-transformed range size as potentially informative predictors of speciation rates. We also used three environmental predictors derived from bioclimatic data. These predictors relate to seasonal climate variation, spatial climate variation and long-term climate variation. To obtain seasonal climate variation we used (*iii*) mean values of temperature seasonality (BIO4) for each range. (*iv*) To estimate levels of spatial environmental variation a species may endure, we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature and precipitation seasonality (BIO4 and BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 **largely** reflects the variation in temperature across a species' range (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). **A taxon’s range size often correlates with speciation and extinction rates [@Rosenzweig_1995; @Castiglione_2017]**, so we controlled for the correlation between environmental spatial variation and species' range sizes --- where larger ranges have larger variation in PC1 --- by using the residuals of a fitted general additive model (GAM; \href{https://ssspeciation.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}) as a predictor. To obtain long-term variation in climates for each species range we take (*v*) the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}). The five predictors of environmental variability are not strongly correlated (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and ``R`` code to generate these predictors can be found within the \href{https://ssspeciation.github.io/SexualSelection_Speciation/}{Supplementary Information}.  \phantomsection

@@

We used BAMM to model the dynamics of speciation and extinction across the **101 phylogenetic trees (one MCC tree and 100 random draws of the posterior)**. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates [@Rabosky_2014]. **These models provide tip-rate estimates of speciation and extinction rate that can be easily used in comparative analyses**. The parameters of the 100 BAMM runs are detailed in full in the \href{https://ssspeciation.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{Supplementary Information}; **briefly, we used a time-variable model with the prior expected number of evolutionary rate shifts set at 100 and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. BAMM models were run independently for the 101 phylogenetic trees for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Important BAMM parameters (log-likelihood and number of rate shifts) reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200; an arbitrary value, above which posterior distributions can often be accurately inferred** (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://ssspeciation.github.io/SexualSelection_Speciation/}{Supplementary Information}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.  \phantomsection

@@

Finally, using the subset of species with an index of male-biased sexual selection, we conducted a phylogenetic path analysis using the ``phylopath R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths between variables unable to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection, and range size to have effects on each other and not just on speciation rate. To minimise path complexity we used temperature seasonality (BIO4) as the single measure for environmental variability, $\lambda_{DR}$ as the single measure of speciation and the tip-rates from the MCC tree. Further details of the path analysis, including our rationale for each path's directions, can be found within the \href{https://ssspeciation.github.io/SexualSelection_Speciation/}{Supplementary Information} along with all other analyses and the relevant ``R`` code to reproduce results.  \phantomsection

@@

We found a significant positive association between the index of male-biased sexual selection (*n* = 2,465) and $\lambda_{DR}$ maximum credibility (MCC) tree ($\beta$ = `r sanitize.numbers(formatC(Subset.MCC.DR.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.DR.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; \autoref{model_results}b). However, this association was not significant for the other two measures of speciation rate ($\lambda_{ND}$: $\beta$ = `r sanitize.numbers(formatC(Subset.MCC.ND.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.ND.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; $\lambda_{BAMM}$: $\beta$ = `r sanitize.numbers(formatC(Subset.MCC.Lambda.top[["coefficients"]][["Sexual_selection_ppca"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(Subset.MCC.Lambda.top)[["Sexual_selection_ppca","p-value"]], digits = 2)`; \autoref{model_results}b). When we took into account phylogenetic uncertainty by running the models using 1,000 trees, the distribution of estimates from PGLS models was similar to the estimate from the MCC tree: among the 1,000 trees there was a positive association between sexual selection and $\lambda_{DR}$ (highest posterior density (HPD) Interval = `r sanitize.numbers(formatC(Subset.hpd.DR.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`), and the distribution skewed towards a positive association between sexual selection and $\lambda_{ND}$ (HPD Interval = `r sanitize.numbers(formatC(Subset.hpd.ND.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`; ) as well as the 100 models using $\lambda_{BAMM}$ (HPD Interval = `r sanitize.numbers(formatC(Subset.hpd.Lambda.top$Sexual_selection_ppca, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`; \href{https://ssspeciation.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Table S15}).

@@

In contrast to male-biased sexual selection, we found no evidence that species with increased sexual dichromatism have higher or lower rates of speciation. Sexual dichromatism showed no association with $\lambda_{DR}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.DR.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.DR.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a, \autoref{Speciation_Figure}), $\lambda_{ND}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.ND.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.ND.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a) or $\lambda_{BAMM}$ ($\beta$ = `r sanitize.numbers(formatC(MCC.Lambda.top[["coefficients"]][["SDi"]], format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, p = `r round(anova(MCC.Lambda.top)[["SDi","p-value"]], digits = 2)`; \autoref{model_results}a). PGLS analyses using sexual dichromatism (*n* = 581) measured by spectrophotometry [@Armenta_2008] yielded results concordant with the full dataset; i.e. no association between sexual dichromatism and speciation  (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S11}). Our results from models based on the MCC tree are largely corroborated by model estimates from PGLS analyses of the rates and correlation structures from 1,000 trees (for $\lambda_{DR}$, $\lambda_{ND}$) and 100 trees for $\lambda_{BAMM}$. The HPD intervals show model estimates are distributed around zero when using complete taxon sampling models and RGB measures of sexual dichromatism ($\lambda_{DR}$: HPD Interval = `r sanitize.numbers(formatC(hpd.DR.top$SDi, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, $\lambda_{ND}$: HPD Interval = `r sanitize.numbers(formatC(hpd.ND.top$SDi, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`, \autoref{model_results}a, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#pgls_model_summary}{Table S8}). For PGLS models using spectrophotometry-based measures of sexual dichromatism, the estimates from the 100 trees in the $\lambda_{DR}$ models are positively skewed (HPD Interval = `r sanitize.numbers(formatC(Armenta.hpd.DR.top$Colour.discriminability, format = "e", digits = 2), type = "latex", math.style.exponents = TRUE)`) but normally distributed around zero for $\lambda_{ND}$ and $\lambda_{BAMM}$ (\href{https://ssspeciation.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Table S12}).

**Our analyses also show that the differences in results between sexual dichromatism and male-biased sexual selection (i.e. association with speciation rates only for the latter) were not due to differences in the size of the datasets used (5,812 species vs. 2,465,** \href{https://ssspeciation.github.io/SexualSelection_Speciation/#testing_the_sexual_dichromatism_dataset_using_the_same_data_points_as_the_sexual_selection_dataset}{Figure S17}). No interaction terms were present in the top models ($\Delta$ AICc > 4) for any measure of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) or sexual selection (RGB values, spectrophotometry and the index of male-biased sexual selection; $\Delta$ AICc > 4; \href{https://ssspeciation.github.io/SexualSelection_Speciation/#pgls_models_on_dr_and_nd}{Table S5}, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#pgls_models_on_bamm_estimates}{Table S6}, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Table S11}, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Table S14}). Thus we found no evidence that the effect of sexual selection on speciation is dependent on our measures of environmental variation or range size. Furthermore, we found no evidence that these environmental factors --- seasonal temperature variation, long-term temperature variation, spatial temperature variation, and Net Primary Productivity (NPP) --- predict speciation independently from sexual dichromatism/selection (\autoref{model_results}, \href{https://ssspeciation.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S11}).  \phantomsection
